Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-05-19 ms.Rnw
use MCAP

Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
In particular, let $\leaves$ be a countable set of labels and let $\part(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
That is, an element $\zeta\in{\part(\leaves)}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
Partition \emph{fineness} defines a partial order on $\part(\leaves)$.
Specifically, for $\zeta,\zeta'\in{\part(\leaves)}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\part(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
An element $b\in{Z_t}$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.
We use the notation $\event{Z}$ to denote the set of times at which $Z$ is discontinuous.
Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
Therefore, $\event{Z}\subseteq{\event{G}}$.

@@

Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in\event{P}$.
The local structure of $P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
For example, if the event in $P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
That is, $Q_u(\eta,\eta')=1$ if and only if
there is a feasible genealogy, $\G=(\T,\Z,\Y)$, and history, $\H$,
and a $t\in{[0,\T]}$ such that,
given $\Gr_\T=\G$ and $\Hr_\T=\H$,
we have $\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
We refer to $Q$ as the \emph{compatibility indicator}.

@@

%% \begin{equation*}
%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
%% \end{equation*}

@@

    foreach (
      p=iter(params,"row"),
      .combine=bind_rows,
      .options.future=list(seed=TRUE)
    ) %dofuture% {
      library(pomp)
      library(phylopomp)
      p |>
        with({
          sirs_tree |>
            sirs_pomp(
              Beta=Beta,gamma=gamma,psi=psi,omega=omega,
              S0=S0,I0=I0,R0=R0,t0=0
            ) |>
            pfilter(Np=Np) |>
            logLik()
        }) -> ll
      bind_cols(p,logLik=ll)
    } -> params
  }
) -> params
@

\begin{figure}
  \begin{center}
    <<sirs3_plot,dependson="sirs1",fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=sirs_tree |>
        plot(points=FALSE,palette=c("#000000"))+
        labs(x="time"),
      B=params |>
        group_by(Beta,gamma,psi,omega,S0,I0,R0,Np) |>
        reframe(
          name=c("logLik","logLik_se"),
          value=logmeanexp(logLik,se=TRUE)
        ) |>
        ungroup() |>
        pivot_wider() |>
        mutate(
          y=logLik,
          ymax=logLik+2*logLik_se,
          ymin=logLik-2*logLik_se
        ) |>
        filter(round(Beta,2)==4) |>
        filter(logLik>max(logLik)-16) |>
        ggplot(
          aes(
            x=gamma,
            group=Np,
            y=y,
            ymin=ymin,
            ymax=ymax
          )
        )+
        geom_errorbar(position="dodge")+
        geom_vline(xintercept=sirs.params$gamma,color="red")+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(gamma)
        )+
        theme_classic()+
        theme(
          legend.position=c(0.5,0.27)
        ),
      labels="AUTO",
      nrow=1,
      rel_widths=c(1,1)
    )
    @
  \end{center}
  \caption{
    \label{fig:sirs_example}
    Likelihood computation for the SIRS model by sequential Monte Carlo.
    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
    (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
    Error bars show $\pm{2}$ s.d.\ of the Monte Carlo estimates.
  }
\end{figure}

@@

This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$;
its deme occupancy function is $n(x)=(E,I)$.
There are five kinds of jumps:
transmission, progression, recovery, waning of immunity, and sampling.
The corresponding marks are $\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$.
\Cref{tab:seirs_model_elements} gives $\alpha_u$, $r_u$, and the event type for each of these marks.

@@

params |>
  group_by(sigma) |>
  summarize(
    sd=sd(logLik),
    logLik=logmeanexp(logLik),
    hi=logLik+2*sd,
    lo=logLik-2*sd
  ) |>
  ungroup() -> stats
@

\begin{figure}
  \begin{center}
    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=seirs_tree |>
        plot(points=TRUE,palette="#000000")+
        labs(x="time"),
      B=params |>
        filter(logLik>max(logLik-16)) |>
        ggplot(aes(x=sigma,y=logLik))+
        geom_point()+
        geom_smooth()+
        geom_vline(xintercept=seirs.params$sigma,color="red")+
        geom_hline(
          yintercept=max(stats$logLik)-
            c(0,0.5*qchisq(p=0.95,df=1)),
          linetype=2
        )+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(sigma)
        )+
        theme_classic(),
      labels="AUTO",
      nrow=1,
      rel_widths=c(3,4)
    )
    @
  \end{center}
  \caption{
    \label{fig:seirs_example}
    Likelihood computation for the SEIRS model by sequential Monte Carlo.
    (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
    $\sigma=\Sexpr{seirs.params$sigma}$,
    $\gamma=\Sexpr{seirs.params$gamma}$,
    $\psi=\Sexpr{seirs.params$psi}$,
    $\omega=\Sexpr{seirs.params$omega}$,
    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs.params,c(S0,E0,I0,R0))})$.
    (B)~Likelihood slice in the $\sigma$-direction.
    Each point represents the estimate of a distinct sequential Monte Carlo computation.
  }
\end{figure}

>>>>

Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
In particular, let $\leaves$ be a countable set of labels and let $\part(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
That is, an element $z\in{\part(\leaves)}$ is a partition of the finite set $\bigcup{z}\subseteq\leaves$.
Partition \emph{fineness} defines a partial order on $\part(\leaves)$.
Specifically, for $z,z'\in{\part(\leaves)}$, we say $z\preceq{z'}$ if and only if for every $b'\in{z'}$ there is $b\in{z}$ such that $b\supseteq{b'}$.
The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\part(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
An element $b\in{Z_t}$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.
We use the notation $\event{Z}$ to denote the set of times at which $Z$ is discontinuous.
Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
Therefore, $\event{Z}\subseteq{\event{G}}$.

@@

Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in\event{P}$.
The local structure of $P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
For example, if the event in $P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
That is, $Q_u(y,y')=1$ if and only if
there is a feasible genealogy, $\G=(\T,\Z,\Y)$, and history, $\H$,
and a $t\in{[0,\T]}$ such that,
given $\Gr_\T=\G$ and $\Hr_\T=\H$,
we have $\U_t=u$, $\Yt_t=y$, and $\Y_t=y'$.
We refer to $Q$ as the \emph{compatibility indicator}.

@@

%% \begin{equation*}
%%   \phi_u(\xi,y,y')\colonequals\BinRatio{n(\xi)}{\ell(y')}{r^u}{s(y,y')}\,Q_u(y,y').
%% \end{equation*}

@@

    foreach (
      p=iter(params,"row"),
      .combine=bind_rows,
      .options.future=list(seed=TRUE)
    ) %dofuture% {
      library(pomp)
      library(phylopomp)
      p |>
        with({
          sirs_tree |>
            sirs_pomp(
              Beta=Beta,gamma=gamma,psi=psi,omega=omega,
              S0=S0,I0=I0,R0=R0,t0=0
            ) |>
            pfilter(Np=Np) |>
            logLik()
        }) -> ll
      bind_cols(p,logLik=ll)
    } -> params
  }
) -> params

params |>
  with(
    mcap(logLik,gamma)
  ) -> mcap
@

\begin{figure}
  \begin{center}
    <<sirs3_plot,dependson="sirs1",fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=sirs_tree |>
        plot(points=FALSE,palette=c("#000000"))+
        labs(x="time"),
      B=params |>
        ggplot(aes(x=gamma,y=logLik))+
        geom_point(alpha=0.2)+
        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
        geom_vline(xintercept=sirs.params$gamma,color="red")+
        geom_vline(xintercept=mcap$ci,linetype=2)+
        geom_hline(
          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
          linetype=2
        )+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(gamma)
        )+
        ## lims(y=c(max(params$logLik)-16,NA))+
        theme_classic()+
        theme(
          legend.position=c(0.5,0.27)
        ),
      labels="AUTO",
      nrow=1,
      rel_widths=c(1,1)
    )
    @
  \end{center}
  \caption{
    \label{fig:sirs_example}
    Likelihood computation for the SIRS model by sequential Monte Carlo.
    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
    (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
    Each point is a distinct Monte Carlo estimate.
    The blue curve is a LOESS smooth; the dashed lines show Monte Carlo-adjusted 95\% confidence intervals \citep{Ionides2017}.
  }
\end{figure}

@@

This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$.
Its deme occupancy function is $n(x)=(E,I)$.
There are five kinds of jumps:
transmission, progression, recovery, waning of immunity, and sampling.
The corresponding marks are $\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$.
\Cref{tab:seirs_model_elements} gives $\alpha_u$, $r_u$, and the event type for each of these marks.

@@

params |>
  with(
    mcap(logLik,sigma,span=0.25)
  ) -> mcap
@

\begin{figure}
  \begin{center}
    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=seirs_tree |>
        plot(points=TRUE,palette="#000000")+
        labs(x="time"),
      B=params |>
        ggplot()+
        geom_point(aes(x=sigma,y=logLik))+
        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
        geom_vline(xintercept=seirs.params$sigma,color="red")+
        geom_vline(xintercept=mcap$ci,linetype=2)+
        geom_hline(
          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
          linetype=2
        )+
        ##        lims(y=c(max(params$logLik)-12,NA))+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(sigma)
        )+
        theme_classic(),
      labels="AUTO",
      nrow=1,
      rel_widths=c(3,4)
    )
    @
  \end{center}
  \caption{
    \label{fig:seirs_example}
    Likelihood computation for the SEIRS model by sequential Monte Carlo.
    (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
    $\sigma=\Sexpr{seirs.params$sigma}$,
    $\gamma=\Sexpr{seirs.params$gamma}$,
    $\psi=\Sexpr{seirs.params$psi}$,
    $\omega=\Sexpr{seirs.params$omega}$,
    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs.params,c(S0,E0,I0,R0))})$.
    (B)~Likelihood slice in the $\sigma$-direction.
    Each point represents the estimate of a distinct sequential Monte Carlo computation.
    The blue curve shows a LOESS smooth;
    the dashed vertical lines enclose the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
  }
\end{figure}
